\Large{ Authors' reply to the comments by M. deâ Michieli Vitturi (Referee \#1) of the manuscript gmd-2017-119} \\ \LARGE \textbf{Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed particle hydrodynamics}

### 2017-09-22 Reply-to-RC1.tex
done proof reading

Yes, dynamic equilibrium means same velocity while thermodynamic equilibrium means same temperature. We treat erupted material (includes volcanic gas and particles) as one phase, so assumption of dynamic equilibrium and thermodynamic equilibrium are made for erupted material. Such assumption is made implicitly when we assume that erupted material behaves like a single phase fluid.
What we intended to say by ``dynamic and thermodynamic equilibrium between air and erupted material" is actually ``dynamic equilibrium and thermodynamic equilibrium between erupted material and air that is entrained into the plume". That is to say, we assume that surrounding air and erupted material will reach an equilibrium status (same velocity and temperature) as soon as they get mixed.

@@

However, the literature reveals that clear transient interfaces do exist in miscible flow. There are experiments \citep {papantoniou1989large} showing a sharp interface between air and plume. \citet{crimaldi2001high} present another experiment with higher resolution that shows an even more detailed and more complicated interface. We postulate that it is  
through the interface (or boundary) that mixing happens. Analogously, \citet{jacobson2008mixing} study has detailed mixing mechanism at a flat interface.

@@

Modification in the manuscripts:
1) Since the exact location of interface is less important for plume modeling when turbulence model is adopted, we changed this paragraph to: 
``Interface tracking is explicit in SPH through capturing of the locations of the particles. Less numerical effort is required for interface construction when we attempt to include the effects of mixing by resolving the detailed interface structure and dynamics of turbulence." and put it as the last item among SPH features.
2) Added a short paragraph (Basically the paragraph above of this response) at the beginning of section 3.8 explaining the motivation for using a turbulence model. 
3) Changing the title of section 3.7 to ``Mass fraction update" we  re-organize this section and focus it more on mass fraction updating. We make a brief remark on interface construction using SPH and mesh-based methods at the end of this section. \\[6pt]

@@

As suggested by the reviewer, we reorganized the order of equations and merged section 3.1 and  section 3.2 into one section. We also added a section for ``Artificial viscosity" before the section on ``governing equation discretization". \\[6pt]

@@

We completely agree with the reviewer that terms for sink associated with phase change should be in the equation of the corresponding phase. For any future extensions of current model, such as the example mentioned by the reviewer, considering phase change of volcanic gas, such sink term should be added to mass conservation equation of volcanic gas. To make the article more readable, we avoid
 discussing new physics models, which are easily understandable for readers working on plume modeling but might cause confusion for readers without such background.
%To avoid confusing readers, we emphasized at the end of section 3.3:  Eq. (42) - Eq. (44) show examples of adding more physics into SPH model based on Euler equations to show easy extensibility of SPH models. The mass source (and sink), drag force effect, and heat exchange effect are not actually added to our model (Eq. (1) - (4)) "
\\[3pt]

@@

If such disequilibrium needs to be considered in future extension of current model, drag force effect should be considered. The form used to represent drag force effects will depend on new assumptions and the new physics model. For example, if the physics model is a four phases model, one phase for atmosphere air, one phase for volcanic gas, one phase for large size solid particles and another for small size ash (maybe immediate equilibrium assumption between air and gas should be made). Drag effects between gas (air) and particles should be considered. But there should be no drag force term within gas phase itself. As for the small size ash, we can assume they behave like single phase fluid (possibly not a reasonable assumption) then there should not drag force effect within the ash phase. Otherwise. drag effect within fine ash phase need to be considered.

We changed the original sentence ``The drag force term should show up only when dynamics disequilibrium between different phases is considered" to ``The drag force term should show up when dynamics disequilibrium between different phases is considered. By deleting the word only we changed original necessary and sufficient statement to a sufficient statement. \\[6pt]

@@

We replaced particle disorder with irregular distribution of particles \\[6pt]

@@

We added explanations and references on tensile instability: ``Tests of the standard SPH method indicate an instability in the tensile regime, while the calculations are stable in compression.  A simple test calculation exhibiting the instability involves a body which is subjected to an uniform initial stress, either compressive or tensile. If the initial stress is tensile, a very small velocity perturbation on a single particle can lead to particles clumping together, forming large voids and seriously corrupting density distribution. But if the initial stress is compressive, the small velocity perturbation on a single particle can not lead to any changes in particle distribution. See paper by \citet{swegle1995smoothed} for more details." \\[3pt]

@@

The sentence is misleading. Sorry for that. Here is the logic: The new approximation of function $A$, Eq. (55), should be consistent with original formulation, Eq. (28). Compare Eq. (55) and Eq. (28), if the denominator is 1, then these two equations will be the same. so it implies that the denominator in Eq. (55) should be close to 1. 
We completely agree with you that the normalization condition originally come from definition of kernel. By the statement ``implies that $...=1$" we were actually trying to show a  consistent connection between this new approximation formulation and properties of the kernel function. In other words, in the derivation based on Taylor series expansion, we never take the property of the kernel into account but finally reach to an equation that is consistent with a property of kernel function.

@@

\citet{price2008modelling} believes that the mixing issue is due to the fact that entropy is discontinuous at the boundaries while density is continuous. He found that adding thermal conductivity at boundaries can improve mixing in SPH as thermal conductivity can smooth the entropy. The thermal conductivity (heat transfer) due to turbulence is considered in our model,  what is more is that the thermal conductivity coefficient is much larger at the interfaces due to large shearing effect at the interface. This can definitely help to relieve the ``mixing challenge" if Price is correct. While, we were not motivated by Price's paper to adding a turbulent heat transfer term into our model -- we were motivatd by the need to have a turbulent heat transfer term for compressible flow in which energy conservation equation is coupled with momentum conservation and mass conservation equation. It appears we were lucky and ``adopted" Price's method.
\citet{wadsley2008treatment} and \citet{ritchie2001multiphase} made similar arguments as Price. \citet{borgani2012hydrodynamic} believes that the ``mixing challenge" of SPH is due to its poor ability of capturing contact discontinuity. They show that using GSPH can avoid appearance of spurious pressure force and help to to follow the Kelvin instability.
Since people still have different opinions on the sources of ``mixing issue", it would be interesting to see the connections between these different opinions. But such work is obviously out of our scope in this paper. These remedies of traditional SPH we adopted and extended in our model helped us to overcome or at least relieve the ``mixing challenge".

@@

In our current model, the ground is assumed to be a flat ground. Since the topography is pretty simple, we did not experience any trouble. it has been shown in some other papers that it works as well for more complex topography \citep{kumar2013parallel}. One potential issue for complicated topography might be deployment of wall ghost particles. For flat boundary, we simply deploy several layers of ghost particles with equal interval and each ghost particle has the same mass. It has not been investigated yet how different deployment of wall ghost particle will influence the accuracy of imposed boundary conditions. A more basic question is how many degree of freedom should we allow for ghost particle deployment: only enough to allow changing of particle position  or enough to allow using different particle mass size?

@@

We have two questions regarding your suggestion: 
1) When we assign pressure value to these static pressure ghost particles, the decrease of pressure due to increase of height ($h$) has already been taken into account. That is to say, pressure of our pressure ghost particles is consistent with meteorological data when establishing the initial condition. Should we still deduce $\rho g h$ from pressure?  
2) Should the total pressure equals to a pressure \textbf{plus} kinetic energy, or minus kinetic energy? Similar questions arise for definition of $p_{\rho}$. Should $\rho g h$ be added to or deduced from pressure? Because total pressure in other fields, such as thermal dynamic analysis of flowing process in engines are defined as pressure plus kinetic energy. %Just to make sure.

@@

Since JPUE is not the phenomena that we aim to model simulation of that is not validation (validation is to make sure simulations are consistent with reality for the application of interest). Meanwhile, we compared our results against experimental results. We might call such comparison (simulation against measurements) as verification in a more general sense (?).

@@

Ritchie's paper proposed two remedies of standard SPH: 1) Keep a density-weighted quantity constant. This might prevent the smoothing length changing in a proper way. One potential issue with such method is that there is no guarantee of enough number of neighbor particles when cavity forms in the fluid field. Probably his method could help us cure the disparity but might also introduce other problems. 2) The second remedy he proposed is assuming a constant pressure within the kernel (not sure whether such assumption can reflect real physics or not). The density is then updated based on equation of state. One good property of density updating equation in standard SPH  adopted in our model is that the total mass is strictly conserved. All other alternative methods for density updating, for example, density updating based on mass conservation PDE, are not able to be not able to guarantee strict conservation of total mass. I did not see any investigation in that paper about how well does his method conserve mass.

@@

We treat all erupted material including particles of different size and volcanic gas as a single phase. So our model is not able to make use of such detailed information. In our model, only the mass fraction of total solid particles is used, and dose not treat particles of different size separately. That's why we did not mention anything about detailed particle size distribution in the input parameter.
 Details about assumptions on which our model relies can be found in section 2. These assumptions are also briefly summarized in abstract. To avoid confusing readers, we added emphasis on which portion of Costa's data are used in our simulation. \\[3pt]

@@

We use a static pressure boundary condition (pressure ghost particles have zero velocity) and do not consider the effect of wind field. Even though wind field data and specific humidity are also in that Fig. 1B in \citet{costa2016results}, we only used density, temperature and pressure but did not use wind velocity and specific humidity in our simulation. 
We added clarifications on this in the revised manuscripts. \\[3pt]

@@

Visualization of the simulated plume is added.

\textit{Page 30, line 1. The averaging technique of Cerminara et al. is defined for grid-based eulerian models. It would be interesting to know the details of the implementation of the technique for the SPH code (perhaps in an appendix).}
\textit{Page 30, lines 1-4.  As particles distribute in a disordered manner in the space in SPH simulation results. We first project simulation results (on disordered particles) onto a pre-defined grid before doing time average and spatial integration. The project method is the basic SPH kernel based interpolation. These lines are confused, please check.}

For these two questions above. We added an appendix describing the posting process in detail. Then we replaced the sentence ``As particles distribute in a disordered manner in the space in SPH
simulation results. We first project simulation results (on disordered particles) onto a pre-defined grid before doing time average and spatial integration. The projection method is the basic SPH kernel based interpolation with  particles distributed irregularly in the space in SPH simulation results. We need to project simulation results (on irregular particles) onto a pre-defined grid before doing time average and spatial integration. See appendix A for more details of post processing."\\[3pt]

@@

Performance benchmarking of our solver is reported in section 3.10. More details have been published in a separate paper. We added an reference to the paper.  As the simulation of Pinatubo reported here was done on computational cluster on CCR in a `` \textbf{non-exclusive}" manner, the computational performance given by a statement like ``spend XXX hours on XXX cpus" might be misleading.

To give an intuitive impression of the ``computational performance" of our solver: it took around 12 days on 240 cores for the Pinatubo simulation. The smoothing length is 170m, $CLF=0.2$, the total duration of simulation is 550 s after eruption. The size of computational domain is $[-40800, 40800] \times [-40800, 40800] \times [1500, 55000] $;
%Even though, we added a paragraph briefly describing the total simulation time that we spend on this volcanic plume simulation.

@@

Here is a summary on major adjustments made in the revised manuscript: 
\begin{itemize}
\item In section 3, moved contents in section 3.2 right behind Eq. (26) and merged section 3.1 and 3.2  into one section. Moved section 3.4 before 3.3. Changed the tile of section 3.7 and adjusted the content in that section.
\item Added a new section in section 4 showing some 1D shock tube tests against analytical results.
\item In section 4, added 4 figures showing mass fraction and velocity of simulated volcanic plume.
\item Added an Appendix on post processing of SPH simulation results.
\end{itemize}

>>>>

Yes, dynamic equilibrium means same velocity while thermodynamic equilibrium means same temperature. We treat erupted material (includes volcanic gas and particles) as one phase, so assumption of dynamic equilibrium and thermodynamic equilibrium are made for erupted material. Such assumption is made implicitly when we assume that erupted material is well mixed and behaves like a single phase fluid.
What we intended to say by ``dynamic and thermodynamic equilibrium between air and erupted material" is actually ``dynamic equilibrium and thermodynamic equilibrium between erupted material and air that is entrained into the plume". That is to say, we assume that surrounding air and erupted material will reach an equilibrium status (same velocity and temperature) as soon as they get mixed.

@@

However, the literature reveals that clear transient interfaces do exist in miscible flow. There are experiments \citep {papantoniou1989large} showing a sharp interface between air and plume. \citet{crimaldi2001high} present another experiment with higher resolution that shows an even more detailed and more complicated interface. We postulate that it is  
through the interface (or boundary) that mixing happens. Analogously, \citet{jacobson2008mixing} studied detailed mixing mechanism at a flat interface.

@@

Modification in the manuscripts:
1) Since the exact location of interface is less important for plume modeling when turbulence model is adopted, we changed this paragraph to: 
``Interface tracking is explicit in SPH through capturing of the locations of the particles. Less numerical effort is required for interface construction when we attempt to include the effects of mixing by resolving the detailed interface structure and dynamics of turbulence." and put it as the last item among SPH features.
2) Added a short paragraph (Basically the paragraph above of this response) at the beginning of section 3.8 explaining the motivation for using a turbulence model. 
3) Changed the title of section 3.7 to ``Mass fraction update". We re-organized this section and focused it more on mass fraction updating. We make a brief remark on interface construction using SPH and mesh-based methods at the end of this section. \\[6pt]

@@

As suggested by the reviewer, we reorganized the order of equations and merged section 3.1 and  section 3.2 into one section. We also moved the section for ``Artificial viscosity" (section 3.4 in original version) before the section on ``governing equation discretization" (section 3.3 in original version). \\[6pt]

@@

We completely agree with the reviewer that terms for sink associated with phase change should be in the equation of the corresponding phase. For future extensions of current model, such as the example mentioned by the reviewer, considering phase change of volcanic gas, such sink term should be added to mass conservation equation of volcanic gas. To make the article more readable, we avoid
 discussing new physics models, which are easily understandable for readers working on plume modeling but might cause confusion for readers without such background.
%To avoid confusing readers, we emphasized at the end of section 3.3:  Eq. (42) - Eq. (44) show examples of adding more physics into SPH model based on Euler equations to show easy extensibility of SPH models. The mass source (and sink), drag force effect, and heat exchange effect are not actually added to our model (Eq. (1) - (4)) "
\\[3pt]

@@

If such disequilibrium needs to be considered in future extension of current model, drag force effect should be considered. The form used to represent drag force effects will depend on new assumptions and the new physics model. For example, if the physics model is a four phases model, one phase for atmosphere air, one phase for volcanic gas, one phase for large size solid particles and another for small size ash (maybe immediate equilibrium assumption between air and gas should be made). Drag effects between gas (air) and particles should be considered. As for the small size ash, we can assume they behave like single phase fluid (possibly not a reasonable assumption) then there should not drag force effect within the ash phase. Otherwise. drag effect within fine ash phase need to be considered.

We changed the original sentence ``The drag force term should show up only when dynamics disequilibrium between different phases is considered" to ``The drag force term should show up when dynamics disequilibrium between different phases is considered. By deleting the word ``only",  we changed original necessary and sufficient statement to a sufficient statement. \\[6pt]

@@

We replaced ``particle disorder" with ``irregular distribution of particles" \\[6pt]

@@

We added explanations and references on tensile instability: ``Tests of the standard SPH method indicate an instability in the tensile regime, while the calculations are stable in compression.  A simple test calculation exhibiting the instability involves a body which is subject to an uniform initial stress, either compressive or tensile. If the initial stress is tensile, a very small velocity perturbation on a single particle can lead to particles clumping together, forming large voids and seriously corrupting density distribution. But if the initial stress is compressive, the small velocity perturbation on a single particle can not lead to any changes in particle distribution. See paper by \citet{swegle1995smoothed} for more details." \\[3pt]

@@

The sentence is misleading. Sorry for that. Here is the logic: The new approximation of function $A$, Eq. (55), should be consistent with original formulation, Eq. (28). Compare Eq. (55) and Eq. (28), if the denominator is 1, then these two equations will be the same. so it implies that the denominator in Eq. (55) should be close to 1. 
We completely agree with you that the normalization condition originally comes from property of kernel function. By the statement ``implies that $...=1$" we were actually trying to show a  consistent connection between this new approximation formulation and properties of the kernel function. In other words, in the derivation based on Taylor series expansion, we never take the property of the kernel into account but finally reach to an equation that is consistent with a property of kernel function.

@@

\citet{price2008modelling} believes that the mixing issue is due to the fact that entropy is discontinuous at the boundaries while density is continuous. He found that adding thermal conductivity at boundaries can improve mixing in SPH as thermal conductivity can smooth the entropy. The thermal conductivity (heat transfer) due to turbulence is considered in our model,  what's more, the thermal conductivity coefficient is much larger at the interfaces due to large shearing effect at the interface. This can definitely help to mitigate the ``mixing challenge" if Price is correct. While, we were not motivated by Price's paper to adding a turbulent heat transfer term into our model -- we were motivatd by the need to have a turbulent heat transfer term for compressible flow in which energy conservation equation is coupled with momentum conservation and mass conservation equation. It appears we were lucky and ``adopted" Price's method.
\citet{wadsley2008treatment} and \citet{ritchie2001multiphase} made similar arguments as Price. \citet{borgani2012hydrodynamic} believes that the ``mixing challenge" of SPH is due to its poor ability of capturing contact discontinuity. They show that using GSPH can avoid appearance of spurious pressure force and help to to follow the Kelvin instability.
Since people still have different opinions on the sources of ``mixing issue", it would be interesting to see the connections between these different opinions. But such work is obviously out of our scope in this paper. These remedies of traditional SPH we adopted and extended in our model helped us to overcome or at least relieve the ``mixing challenge".

@@

In our current model, the ground is assumed to be a flat ground. Since the topography is pretty simple, we did not experience any trouble. It has been shown in some other papers that the approach for imposing wall boundary conditions also works as well for more complex topography \citep{kumar2013parallel}. One potential issue for complicated topography might be deployment of wall ghost particles. For flat boundary, we simply deploy several layers of ghost particles with equal interval and each ghost particle has the same mass. It has not been investigated yet how different deployment of wall ghost particle will influence the accuracy of imposed boundary conditions. A more basic question is how many degree of freedom should we allow for ghost particle deployment: only enough to allow changing of particle position  or enough to allow using different particle mass size?

@@

We have two questions regarding your suggestion: 
1) When we assign pressure value to these static pressure ghost particles, the decrease of pressure due to increase of height ($h$) has already been taken into account. That is to say, pressure of our pressure ghost particles is consistent with meteorological data when establishing the initial condition. Should we still deduce $\rho g h$ from pressure?  
2) Should the total pressure equals to a pressure \textbf{plus} kinetic energy, or minus kinetic energy? Similar questions arise for definition of $p_{\rho}$. Should $\rho g h$ be added to or deduced from pressure? Because total pressure in other fields, such as thermal dynamic analysis of flowing process in engines is defined as pressure plus kinetic energy. %Just to make sure.

@@

Since JPUE is not the phenomena that we aim to model. Simulation of JPUE should not be validation (validation is to make sure simulations are consistent with reality for the application of interest). Meanwhile, we compared our results against experimental results. We might call such comparison (simulation against measurements) as verification in a more general sense (?).

@@

Ritchie's paper proposed two remedies of standard SPH: 1) Keep a density-weighted quantity constant. This might prevent the smoothing length changing in a proper way. One potential issue with such method is that there is no guarantee of enough number of neighbor particles when cavity forms in the fluid field. Probably his method could help us cure the disparity but might also introduce other problems. 2) The second remedy he proposed is assuming a constant pressure within the kernel (not sure whether such assumption can reflect real physics or not). The density is then updated based on equation of state. One good property of density updating equation in standard SPH  adopted in our model is that the total mass is strictly conserved. All other alternative methods for density updating, for example, density updating based on mass conservation PDE, are not able to guarantee strict conservation of total mass. I did not see any investigation in that paper about how well does his method conserve mass.

@@

We treat all erupted material including particles of different size and volcanic gas as a single phase. So our model is not able to make use of such detailed information. In our model, only the mass fraction of total solid particles is used, and dose not treat particles of different size separately. That's why we did not mention anything about detailed particle size distribution in the table for input parameters.
 Details about assumptions on which our model relies can be found in section 2. These assumptions are also briefly summarized in abstract. To avoid confusing readers, we added emphasis on which portion of Costa's data are used in our simulation. \\[3pt]

@@

We use a static pressure boundary condition (pressure ghost particles have zero velocity) and do not consider the effect of wind field. Even though wind field data and specific humidity are also provided in Fig. 1B in the paper by \citet{costa2016results}, we only used density, temperature and pressure and did not use wind velocity and specific humidity in our simulation. 
We added clarifications on this in the revised manuscripts. \\[3pt]

@@

Visualization of the simulated plume is added \\[3pt]

\textit{Page 30, line 1. The averaging technique of Cerminara et al. is defined for grid-based eulerian models. It would be interesting to know the details of the implementation of the technique for the SPH code (perhaps in an appendix).}

\textit{Page 30, lines 1-4.  As particles distribute in a disordered manner in the space in SPH simulation results. We first project simulation results (on disordered particles) onto a pre-defined grid before doing time average and spatial integration. The project method is the basic SPH kernel based interpolation. These lines are confused, please check.}

For these two questions above. We added an appendix describing the posting process in detail. Then we replaced the sentence ``As particles distribute in a disordered manner in the space in SPH
simulation results. We first project simulation results (on disordered particles) onto a pre-defined grid before doing time average and spatial integration. The projection method is the basic SPH kernel based interpolation" by a new sentence  ``As particles distribute irregularly in the space in SPH simulation results. We need to project simulation results (on irregular particles) onto a pre-defined grid before doing time average and spatial integration. See appendix A for more details of post processing." \\[3pt]

@@

Performance benchmarking of our solver is reported in section 3.10 separately. More details have been published in a separate paper. We added an reference to the paper.  As the simulation of Pinatubo reported here was done on computational cluster sharing computational nodes with other users, the computational performance given by a statement like ``spend XXX hours on XXX cpus" might be misleading.

To give an intuitive impression of the ``computational performance" of our solver: it took around 15 days on 240 cores for the Pinatubo simulation. The smoothing length is 170m, $CLF=0.2$, the total duration of simulation is 550 s after eruption. The size of computational domain is $[-40800, 40800] \times [-40800, 40800] \times [1500, 55000]$;
%Even though, we added a paragraph briefly describing the total simulation time that we spend on this volcanic plume simulation.

@@

Here is a summary on major adjustments made in the revised manuscript: 
\begin{itemize}
\item In section 3, moved contents in section 3.2 right behind Eq. (26) and merged section 3.1 and 3.2  into one section. Moved section 3.4 before 3.3. Changed the tile of section 3.7 and adjusted the content in that section.
\item Added a new subsection in section 4 showing some 1D shock tube tests against analytical results.
\item In section 4, added 4 figures showing mass fraction and velocity of simulated volcanic plume.
\item Added an Appendix on post processing of SPH simulation results.
\end{itemize}
